---
title: "Systematic review and meta-analysis of Real-World Evidence"
authors:   
  - name: Dimitris Mavridis
    affiliations:
      - ref: ioannina
  - name: Thomas Debray
    orcid: 0000-0002-1790-2719
    affiliations:
      - ref: smartdas
affiliations:
  - id: smartdas
    name: Smart Data Analysis and Statistics B.V.
    city: Utrecht
  - id: ioannina
    name: University of Ioannina
    city: Ioannina
format:
  html:
    toc: true
    number-sections: true
execute:
  cache: true
bibliography: 'https://api.citedrive.com/bib/0d25b38b-db8f-43c4-b934-f4e2f3bd655a/references.bib?x=eyJpZCI6ICIwZDI1YjM4Yi1kYjhmLTQzYzQtYjkzNC1mNGUyZjNiZDY1NWEiLCAidXNlciI6ICIyNTA2IiwgInNpZ25hdHVyZSI6ICI0MGFkYjZhMzYyYWE5Y2U0MjQ2NWE2ZTQzNjlhMWY3NTk5MzhhNzUxZDNjYWIxNDlmYjM4NDgwOTYzMzY5YzFlIn0=/bibliography.bib'
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(kableExtra)
```

## Introduction

We first load the required packages

```{r}
#| warning: false
library(dplyr)
library(netmeta)
```



## Advanced Therapies for Ulcerative Colitis
In this example, we consider a systematic literature review of Phase 3 
randomized controlled trials investigating the following advanced therapies: 
infliximab, adalimumab, vedolizumab, golimumab, tofacitinib, ustekinumab, 
filgotinib, ozanimod, and upadacitinib [@panaccione_efficacy_2023]. This review 
included 48 RCTs, from which 23 were found eligible for inclusion in a 
network meta-analysis. 

The included RCT populations were largely comparable in their baseline
characteristics, though some heterogeneity was noted in weight, 
disease duration, extent of disease, and concomitant medications. 
A risk of bias assessment showed a low risk of bias for all included RCTs, 
which were all industry sponsored.

### Example Dataset
We here focus on the synthesis of 18 trials that contributed efficacy data for 
induction in bio-naive populations. The following FDA- and/or EMA-approved 
biologic or SMD doses were investigated: 

* Adalimumab subcutaneous 160 mg at week 0, 80 mg at week 2, and 40 mg at week 4 (`ADA160/80`)
* Infliximab intravenous 5 mg/kg (`INF5`) at weeks 0, 2, and 6 then every 8 weeks
* Infliximab intravenous 10 mg/kg (`INF10`) at weeks 0, 2, and 6 then every 8 weeks
* Filgotinib oral 100 mg once daily (`FIL100`)
* Filgotinib oral 200 mg once daily (`FIL200`)
* Golimumab subcutaneous 200 mg at week 0 and 100 mg at week 2 (`GOL200/100`)
* Ozanimod oral 0.23 mg once daily for 4 days, 0.46 mg once daily for 3 days, then 0.92 mg once daily (`OZA0.92`)
* Tofacitinib oral 10 mg twice daily for 8 weeks (`TOF10`)
* Upadacitinib oral 45 mg once daily for 8 weeks (`UPA45`)
* Ustekinumab intravenous 6 mg/kg at week 0 (`UST6`)
* Vedolizumab intravenous 300 mg at weeks 0, 2, and 6 (`VED300`)

The reference treatment is placebo (`PBO`).


```{r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: Efficacy outcomes (i.e., clinical remission) data of induction bio-naïve populations
# Efficacy outcomes data of induction bio-naïve populations
studlab <- c("ACT-1", "ACT-1", "ACT-1", "ACT-2", "ACT-2", "ACT-2", "GEMINI 1",
             "GEMINI 1", "Japic CTI-060298", "Japic CTI-060298", "Jiang 2015", 
             "Jiang 2015", "M10-447", "M10-447", "NCT01551290", "NCT01551290",
             "NCT02039505", "NCT02039505", "OCTAVE 1", "OCTAVE 1", "OCTAVE 2",
             "OCTAVE 2", "PURSUIT-SC", "PURSUIT-SC", "SELECTION", "SELECTION",
             "SELECTION", "TRUE NORTH", "TRUE NORTH", "U-ACCOMPLISH", 
             "U-ACCOMPLISH", "U-ACHIEVE Study 2", "U-ACHIEVE Study 2", 
             "ULTRA-1", "ULTRA-1", "ULTRA-2", "ULTRA-2", "UNIFI", "UNIFI")

treat <- c("INF10", "INF5", "PBO", "INF10", "INF5", "PBO", "VED300", "PBO", 
           "INF5", "PBO", "INF5", "PBO", "ADA160/80", "PBO", "INF5", "PBO", 
           "VED300", "PBO", "TOF10", "PBO", "TOF10", "PBO", "GOL200/100", 
           "PBO", "FIL100", "FIL200", "PBO", "OZA0.92", "PBO", "UPA45", "PBO",
           "UPA45", "PBO", "ADA160/80", "PBO", "ADA160/80", "PBO", "UST6",
           "PBO")


event <- c(39, 47, 18, 33, 41, 7, 30, 5, 21, 11, 22, 9, 9, 11, 11, 5, 22,
           6, 56, 9, 43, 4, 45, 16, 47, 60, 17, 66, 10, 54, 3, 41, 4, 24,
           12, 32, 16, 27, 15)

n <- c(122, 121, 121, 120, 121, 123, 130, 76, 104, 104, 41, 41, 90, 96, 50,
       49, 79, 41, 222, 57, 195, 47, 253, 251, 277, 245, 137, 299, 151, 166,
       81, 145, 72, 130, 130, 150, 145, 147, 151)

UlcerativeColitis <- pairwise(treat=treat,event=event,n=n,studlab=studlab, sm="OR")

UlcerativeColitis %>% select(studlab, treat1, treat2, event1, n1, event2, n2) %>% 
  kable(
      align = "lrrrrrr",
      booktabs = TRUE,
      longtable = TRUE,
      linesep = ""
    ) %>%
    kableExtra::kable_styling(
      latex_options = c("repeat_header"),
    )
```

The corresponding network is displayed below:

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-cap: Evidence network of 18 trials that contributed efficacy data for induction in bio-naive populations

NMA.uc <- netmeta(TE = TE, seTE = seTE, treat1 = treat1, treat2 = treat2,
                  studlab = studlab, data = UlcerativeColitis, sm = "OR", 
                  ref = "PBO", comb.random = TRUE)

netgraph(NMA.uc)
```

### Network meta-analysis
Below, we conduct a random effects network meta-analysis of the reported study 
effects (expressed as odds ratio) and consider placebo (`treat = "PBO"`) as 
the control treatment.

```{r}
#| eval: false
NMA.uc <- netmeta(TE = TE, seTE = seTE, treat1 = treat1, treat2 = treat2,
                  studlab = studlab, data = UlcerativeColitis, sm = "OR", 
                  ref = "PBO", comb.random = TRUE)
```

```{r}
#| echo: false
#| message: false
#| warning: false
plot(NMA.uc)
```

## COVID-19

### Example Dataset
```{r}
#| echo: false
#| message: false
#| warning: false
studlab <- c(rep("Tomazini et al",2), rep("Horby et al",2), 
             rep("Biegel et al",2), rep("Wang et al",2),
             rep("Spinner et al",2),rep("Solidarity trial - Remdesivir",2),
             rep("Hermine et al",2),rep("Salvarini et al",2),
             rep("Stone et al",2), rep("Rosas et al",2), rep("Salama et al",2),
             rep("Veiga et al",2), rep("Ulrich et al",2),
             rep("Lyngbakken et al",2), rep("Solidarity trial - HCQ",2), 
             rep("Abd-Elsalam et al",2),rep("Cavalcanti et al",2),
             rep("Recovery group",2), rep("Solidarity trial-LR",2), 
             rep("Cao et al",2), rep("Davoudi-Monfared",2), 
             rep("Solidarity trial - Interferon",2),rep("Rohmani",2))

treat <- c("DEXA", "STD", "DEXA", "STD", "REM", "STD", "REM", "STD", "REM", 
           "STD", "REM", "STD", "TOCI", "STD", "TOCI", "STD", "TOCI", "STD",
           "TOCI", "STD", "TOCI", "STD", "TOCI", "STD", "HCQ", "STD", "HCQ", 
           "STD", "HCQ", "STD", "HCQ", "STD", "HCQ", "STD", "LOP+RIT",  
           "STD", "LOP+RIT", "STD", "LOP+RIT", "STD", "INT-B", "STD", "INT-B",
           "STD", "INT-B", "STD")    

event=c(85,91,482,1110,59,77,22,10,5,4,301,303,7,8,2,1,9,3,58,28,26,11,14,6,13,6,1,1,104,84,6,5,1,1,374,767,148,146,19,25,8,15,243,216,2,6)
n=c(151,148,2104,4321,541,521,158,78,396,200,2743,2708,63,67,60,66,161,82,294,144,249,128,65,64,67,61,27,26,947,906,97,97,221,227,1616,3424,1399,1372,99,100,42,39,2050,2050,33,33)
```

We have the following treatments: standard care (`treat = "STD"`), ...
```{r}
#| echo: false
#| message: false
#| warning: false

covid <- pairwise(treat=treat,event=event,n=n,studlab=studlab, sm="OR")

covid %>% select(studlab, treat1, treat2, event1, n1, event2, n2) %>% 
  kable(
      align = "lrrrrrr",
      booktabs = TRUE,
      longtable = TRUE,
      linesep = ""
    ) %>%
    kableExtra::kable_styling(
      latex_options = c("repeat_header"),
    )
```

### Network meta-analysis
Below, we conduct a random effects network meta-analysis where we consider 
standard care (`treat = "STD"`) as the control treatment.

```{r}
NMA.covid <- netmeta(TE = TE, seTE = seTE, treat1 = treat1, treat2 = treat2,
                     studlab = studlab, data = covid, sm = "OR", ref = "STD",
                     comb.random = TRUE)
summary(NMA.covid)
```

