---
title: "Dealing with irregular and informative visits"
author: "Thomas Debray"
date: "This report was generated on `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
source("Chapter 10 - Dealing with irregular and informative visits/sim.r")
source("Chapter 10 - Dealing with irregular and informative visits/fig_functions.r")
source("Chapter 10 - Dealing with irregular and informative visits/mlmi.r")

```

# Introduction

We first load the required packages

```{r}
library(dplyr)
library(broom)
library(ggplot2)
library(mice)
```

# Example dataset

In this example dataset, we have a discrete outcome `y` that is affected by its baseline value `edss`, age, sex, and the treatment duration `time`.
```{r}
set.seed(9843626)

dataset  <- sim_data_EDSS(npatients = 500, #10000, # 2000, # previous 50000
                          ncenters = 10,
                          follow_up = 12*5, # Total follow-up (number of months)
                          sd_a_t = 0.5,   # DGM - Within-visit variation in EDSS scores
                          baseline_EDSS = 1.3295,    # DGM - Mean baseline EDDS score
                          sd_alpha_ij = 1.46,    # DGM - Between-subject variation in baseline EDSS
                          sd_beta1_j = 0.20,    # DGM - Between-site variation in baseline EDSS
                          mean_age = 42.41,
                          sd_age = 10.53,
                          min_age = 18,
                          beta_age = 0.05, # DGM - prognostic effect of age
                          beta_t = 0.014,  # DGM - prognostic effect of time
                          beta_t2 = 0,    # DGM - prognostic effect of time squared
                          delta_xt = 0, # DGM - interaction treatment time
                          delta_xt2 = 0, # 0.0005    # DGM - interaction treatment time2
                          p_female = 0.75, 
                          beta_female = -0.2 ,  ## DGM - prognostic effect of male sex
                          delta_xf = 0,      ## DGM - interaction sex treatment       
                          rho = 0.8,             # DGM - autocorrelation of between alpha_tij
                          corFUN = corAR1,       # DGM - correlation structure of the latent EDSS scores
                          tx_alloc_FUN = treatment_alloc_confounding_v2 ) ## or treatment_alloc_randomized
```

We remove `y` according to the informative visit process that depends on the received treatment, gender, and age.

```{r}
dataset_visit <- censor_visits_a5(dataset) %>% select(-y) 
```

In the censored data, a total of `r sum(!is.na(dataset_visit %>% filter(time==60) %>% pull("y_obs")))` out of `r length(unique(dataset_visit$patid))` patients have a visit at `time=60`.



# Estimation of treatment effect
We will estimate the marginal treatment effect at time `time=60`.

## Original data
```{r}
origdat60 <- dataset %>% filter(time == 60)

# Predict probability of treatment allocation
fitps <- glm(x ~ age + sex + edss, family='binomial', data = origdat60)

# Derive the propensity score
origdat60 <- origdat60 %>% mutate(ipt = ifelse(x==1, 1/predict(fitps, type='response'),
                                            1/(1-predict(fitps, type='response'))))

# Estimate 
fit_ref_m <- tidy(lm(y ~ x, weight = ipt, data = origdat60), conf.int = TRUE) 
```


```{r echo = F}
results <- data.frame(method = character(),
                      estimate = numeric(),
                      lci = numeric(),
                      uci = numeric())

results <- results %>% add_row(data.frame(method = "Reference"),
                               estimate = fit_ref_m %>% filter(term == "x") %>% pull(estimate),
                               lci = fit_ref_m %>% filter(term == "x") %>% pull(conf.low),
                               uci = fit_ref_m %>% filter(term == "x") %>% pull(conf.high))

```


## doubly -weighted marginal treatment effect
```{r}
obsdat60 <- dataset_visit %>% mutate(visit = ifelse(is.na(y_obs),0,1)) %>% filter(time == 60)

gamma <- glm(visit ~ x + sex + age + edss, family = 'binomial', data = obsdat60)$coef   

obsdat60 <- obsdat60 %>% mutate(rho_i = 1/exp(gamma["(Intercept)"] +
                                                          gamma["x"]*x +
                                                          gamma["sex"]*sex +
                                                          gamma["age"]*age))

# Predict probability of treatment allocation
fitps <- glm(x ~ age + sex + edss, family='binomial', data = obsdat60)

# Derive the propensity score
obsdat60 <- obsdat60 %>% mutate(ipt = ifelse(x==1, 1/predict(fitps, type='response'),
                                            1/(1-predict(fitps, type='response'))))


fit_w <- tidy(lm(y_obs ~ x, weights = ipt*rho_i, data = obsdat60), conf.int = TRUE)
```

```{r echo = F}
results <- results %>% add_row(data.frame(method = "Doubly weighted"),
                               estimate = fit_w %>% filter(term == "x") %>% pull(estimate),
                               lci = fit_w %>% filter(term == "x") %>% pull(conf.low),
                               uci = fit_w %>% filter(term == "x") %>% pull(conf.high))
```

## Multilevel multiple imputation

We impute the entire vector of `y_obs` for all 61 potential visits and generate 10 imputed datasets. Note: `mlmi` currently does not support imputation of treatment-covariate interaction terms.

```{r, eval = FALSE}
dataset_visit <- dataset_visit %>% mutate(timex = time*x)

method <- rep("", ncol(dataset_visit))
names(method) <- colnames(dataset_visit)
method["y_obs"] <- "mlmi"

predM <- matrix(0, nrow = ncol(dataset_visit), ncol = ncol(dataset_visit))
colnames(predM) <- rownames(predM) <- colnames(dataset_visit)

predM["y_obs", c("x","age","sex","edss","time","timex")] <- 1
predM["y_obs", "time"] <- 6
predM["y_obs", "patid"] <- -3
predM["y_obs", "centerid"] <- -2

imp <- mice(dataset_visit, method = method, predictorMatrix = predM, 
            maxit = 1, m = 10, seed = 12345)
```

```{r echo = F}
load("Chapter 10 - Dealing with irregular and informative visits/imp_mlmi.rda")
```

We can now estimate the treatment effect in each imputed dataset

```{r}
Q <- rep(NA, imp$m) # Point estimates
U <- rep(NA, imp$m) # Error variances

for (i in seq(imp$m)) {
  dati <- complete(imp, i) %>% filter(time == 60)
  
  # Predict probability of treatment allocation
  fitps <- glm(x ~ age + sex + edss, family='binomial', data = dati)
  
  # Derive the propensity score
  dati <- dati %>% mutate(ipt = ifelse(x==1, 1/predict(fitps, type='response'),
                                            1/(1-predict(fitps, type='response'))))
  
  # Estimate 
  fit <- tidy(lm(y_obs ~ x, weight = ipt, data = dati), conf.int = TRUE) 
  
  Q[i] <- fit %>% filter(term == "x") %>% pull(estimate)
  U[i] <- (fit %>% filter(term == "x") %>% pull(std.error))**2
}

fit_mlmi <- pool.scalar(Q = Q, U = U)
```

```{r echo = F}
results <- results %>% add_row(data.frame(method = "Multilevel Multiple Imputation"),
                               estimate = fit_mlmi$qbar,
                               lci = NA,
                               uci = NA)
```

# Results
```{r echo = F}
ggplot(results, aes(x=method, y = estimate)) +
  geom_point() + 
  geom_errorbar(aes(ymin = lci, ymax = uci)) +
  ylab("Marginal treatment effect")
```


