---
title: "Individual Participant Data Meta-analysis of clinical trials and real-world data"
authors:   
  - name: Pablo Verde
    affiliations:
      - ref: umcdusseldorf
  - name: Thomas Debray
    orcid: 0000-0002-1790-2719
    affiliations:
      - ref: smartdas
affiliations:
  - id: smartdas
    name: Smart Data Analysis and Statistics B.V.
    city: Utrecht
  - id: umcdusseldorf
    name: Universitätsklinikum Düsseldorf
    city: Düsseldorf
format:
  html:
    toc: true
    number-sections: true
execute:
  cache: true
bibliography: 'https://api.citedrive.com/bib/0d25b38b-db8f-43c4-b934-f4e2f3bd655a/references.bib?x=eyJpZCI6ICIwZDI1YjM4Yi1kYjhmLTQzYzQtYjkzNC1mNGUyZjNiZDY1NWEiLCAidXNlciI6ICIyNTA2IiwgInNpZ25hdHVyZSI6ICI0MGFkYjZhMzYyYWE5Y2U0MjQ2NWE2ZTQzNjlhMWY3NTk5MzhhNzUxZDNjYWIxNDlmYjM4NDgwOTYzMzY5YzFlIn0=/bibliography.bib'
---


```{r}
#| include: false
#| echo: false
#| message: false
#| warning: false

# List of required packages
required_packages <- c("ggplot2", "dplyr", "gridExtra", "mcmcplots", 
                       "grid", "table1",
                       "gridExtra", "GGally")

# Install required packages
for (pkg in required_packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
        install.packages(pkg)
    }
}
library(kableExtra)
library(table1)
```

## Introduction


## Adjuvant therapies for diabetes
In this example, we consider the synthesis of the following data:

* Aggregate data from 35 randomized trials investigating the effectiveness of adjuvant therapies for diabetic patients, and 
* Individual participant data from a prospective cohort study investigating patient and limb survival in patients with diabetic foot ulcers 

We first retrieve the randomized evidence.

```{r}
#| warning: false
#| message: false
#| results: hide
library(dplyr)
library(jarbes)

data("healing")
AD <- healing %>% dplyr::select(y_c, n_c, y_t, n_t, Study)
names(AD) <- c("yc", "nc", "yt", "nt", "Study")
```

The corresponding aggregate data are depicted below:

```{r}
#| echo: false
kable(AD %>% dplyr::select(Study, yc, nc, yt, nt))
```

Subsequently, we retrieve the individual participant data:

```{r}
#| warning: false
data("healingipd")
IPD <- healingipd %>% dplyr::select(healing.without.amp, PAD, neuropathy,
                                    first.ever.lesion, no.continuous.care, 
                                    male, diab.typ2, insulin, HOCHD, 
                                    HOS, CRF, dialysis, DNOAP, smoking.ever, 
                                    diabdur, wagner.class)
```

Briefly, these IPD were obtained from a prospective cohort study enrolling consecutive patients with diabetic foot ulcers (DFUs) and without previous major amputation in a  single diabetes center between June 1998 and December 1999. The baseline characteristics of the study population is summarized below:

```{r}
#| echo: false
dstbl <- healingipd %>% mutate(PAD = factor(PAD , levels = c(1,0), labels = c("Yes", "No")),
                        neuropathy = factor(neuropathy, levels = c(1,0), labels = c("Yes", "No")),
                        first.ever.lesion = factor(first.ever.lesion, levels = c(1,0), labels = c("Yes", "No")),
                        no.continuous.care = factor(no.continuous.care, levels = c(1,0), labels = c("Yes", "No")),
                        gender = ifelse(male == 1, "Male", "Female"),
                        diab.typ2 = factor(diab.typ2, levels = c(1,0), labels = c("Yes", "No")),
                        insulin = factor(insulin, levels = c(1,0), labels = c("Yes", "No")),
                        HOCHD = factor(HOCHD, levels = c(1,0), labels = c("Yes", "No")),
                        HOS = factor(HOS, levels = c(1,0), labels = c("Yes", "No")),
                        CRF = factor(CRF, levels = c(1,0), labels = c("Yes", "No")),
                        dialysis = factor(dialysis, levels = c(1,0), labels = c("Yes", "No")),
                        DNOAP = factor(DNOAP, levels = c(1,0), labels = c("Yes", "No")),
                        smoking.ever = factor(smoking.ever, levels = c(1,0), labels = c("Yes", "No")))

label(dstbl$age) <- "Age"
label(dstbl$PAD) <- "Peripheral arterial disease"
label(dstbl$neuropathy) <- "Neuropathy"
label(dstbl$first.ever.lesion) <- "First ever lesion"
label(dstbl$no.continuous.care) <- "No continuous care"
label(dstbl$gender) <- "Sex"
label(dstbl$diab.typ2) <- "Diabetes type 2"
label(dstbl$insulin) <- "Insulin dependent"
label(dstbl$HOCHD) <- "History of coronary events (CHD)"
label(dstbl$HOS) <- "History of stroke"
label(dstbl$CRF) <- "Charcot foot syndrome"
label(dstbl$dialysis) <- "Dialysis"
label(dstbl$DNOAP) <- "DNOAP"
label(dstbl$smoking.ever) <- "Ever smoker"
label(dstbl$diabdur) <- "Diabetes duration"
label(dstbl$wagner.class) <- "Wagner score"

units(dstbl$age) <- "years"
units(dstbl$diabdur) <- "years"

table1(~ age + diabdur + gender + smoking.ever + diab.typ2 + PAD + neuropathy + first.ever.lesion +
         no.continuous.care  + insulin + HOCHD +
         HOS + CRF + dialysis + DNOAP +  wagner.class, data = dstbl)

```

We prepare the relevant aggregate data, and subsequently conduct a meta-analysis to analyse heterogeneity of the treatment effect as a function of the baseline risk

```{r}
#| warning: false
#| results: hide
m.0 <- metarisk(data = AD)
summary(m.0)
```

```{r}
#| echo: false
summary(m.0)
```


```{r}
OR <- mean(m.0$BUGSoutput$sims.matrix[,"Odds.pool"])
CI95 <- quantile(m.0$BUGSoutput$sims.matrix[,"Odds.pool"], 
                 probs = c(0.025, 0.975))
```

