---
title: "Dealing with missing data"
author: "Johanna Munoz"
date:  "This report was generated on `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(kableExtra)
library(table1)

source("Chapter 04 - Missing Data Imputation/chapter_4.r")
```
# Prepare R environment

```{r}
library(mice)
library(ggmice)
```

# Absence of HTE

## Generating an observational dataset
We can simulate an observational dataset of $N = 3000$ patients as follows:
```{r}
data_noHTE <- generate_data(n = 3000, seed = 1234) 
```

This dataset does not (yet) contain any missing values;

The simulated dataset contains two treatment groups with differences in baseline characteristics. For example, the figure below shows that we have baseline imbalance in age.

```{r, fig.align="center", fig.width=6, fig.height=6, fig.cap="Figure 1:Distribution of confounders and outcome variable", echo = F}
plot_den(datap = data_noHTE,var="age",varlab="Age (years)")  #Age density
```

```{r, echo = F}
# Estimation of treatment effects in the complete data
true_noHTE_ATE <- getest(data = data_noHTE, estimandv = "ATE", Tform = 1, CC = TRUE, analysis = "Full Data")

```

## Generating missing values
Missing values can be generated using the function `getmissdata()`, which considers the following patterns of missingness for the previous number of relapses (`prerelapse_num`):

1. MAR: missingness depends on `age` and `sex`
2. MART:  missingness depends on `age`, `sex` and the treatment variable `treatment`
3. MARTY: missingness depends on `age`, `sex`, `treatment` and the outcome variable `y`
4. MNAR:  missingness depends on `age`, `sex` and `prerelapse_num` 

```{r}
mdata_noHTE <- getmissdata(data_noHTE, "MART")
```

After introducing missing values, we only have complete data for $N=$ `r sum(complete.cases(mdata_noHTE))` patients.

```{r echo = F}
dat <- formatMSdata(mdata_noHTE)

table1(~ age + female + prevDMTefficacy + premedicalcost + numSymptoms + prerelapse_num | treatment, data = dat, caption = "Baseline characteristics of the incomplete dataset.")
```

## Multiple Imputation

We first need to specify how each variable will be imputed
```{r}
# Specify the conditional imputation models
form_y <- list(prevDMTefficacy ~ age+female + years+premedicalcost+numSymptoms+treatment+prerelapse_num+y,
               premedicalcost~age+female + years+prevDMTefficacy+numSymptoms+treatment+prerelapse_num+y,
               numSymptoms ~age+female + premedicalcost+years+prevDMTefficacy+prerelapse_num+treatment+y,
               prerelapse_num ~age+female + premedicalcost+years+prevDMTefficacy+numSymptoms+treatment+y,
               age ~prerelapse_num+female + premedicalcost+years+prevDMTefficacy+numSymptoms+treatment+y)
form_y <- name.formulas(form_y)

imp0 <- mice(mdata_noHTE, form = form_y, maxit = 0)
method <- imp0$method
method["numSymptoms"] <- "pmm"
method["prevDMTefficacy"] <- "pmm"
```

```{r eval = F}
# Generate 5 imputed datasets
imp <- mice(mdata_noHTE, form = form_y, method = method, m = 5, maxit = 100)
```

We can inspect convergence for the imputed variable `prerelapse_num` using a trace plot:
```{r echo = F}
load("Chapter 04 - Missing Data Imputation/imp_noHTE.rda")
```
```{r}
plot_trace(imp, vrb = "prerelapse_num")
```

## Estimation of treatment effect

Analysis methods:

- **Full Data**: The treatment effect is estimated in the original data of $N=3000$ patients.
- **Complete Case Analysis**: The treatment effect is estimated using the data from $N=$ `r sum(complete.cases(mdata_noHTE))` patients without any missing values.
- **Missing Indicator**: The treatment effect is estimated in the incomplete dataset of $N=3000$ patients. The propensity score model includes a missing indicator variable for each incomplete covariate.
- **Multiple Imputation (within method)**: A treatment effect is estimated within each imputed dataset using propensity score analysis. Using Rubin’s rule, the five treatment effects are combined into a single treatment effect. 

```{r echo = F, message = F}
# Complete case analysis
fit_cc <- getest(data = mdata_noHTE, 
                 estimandv = "ATE", # Estimate the ATE or ATT  
                 Tform = 1, # PS model formula
                 CC = TRUE, # use the complete case dataset,
                 analysis = "Complete Case Analysis")

# Missing indicator
fit_MInd <- getest(data = mdata_noHTE, 
                   estimandv = "ATE", # Estimate the ATE or ATT  
                   Tform = 2, # PS model formula
                   CC = FALSE, # use the complete case dataset,
                   analysis = "Missing Indicator")

# Multiple imputation; matching within each imputed dataset
fit_MICEw <- getmicest(data = imp, 
                       estimandv = "ATE", # Estimate the ATE or ATT  
                       Tform = 1, # PS model formula
                       approachv = "within",
                       analysis = "MICE (within method)")


alldata <- rbind(true_noHTE_ATE, fit_cc, fit_MInd, fit_MICEw)

alldata$analysis = factor(alldata$analysis,
                           labels = c("Full Data", "Complete Case Analysis", 
                                      "Missing Indicator", "MICE (within method)"),
                           levels = c("Full Data", "Complete Case Analysis", 
                                      "Missing Indicator", "MICE (within method)"))

ggplot(alldata,aes(x= analysis, y= estimate, col = analysis)) +
  geom_point(shape = 1,
             size  = 1) +
  geom_errorbar(aes(ymin  = LCI,
                    ymax  = UCI),
                width = 0.2,
                size  = 0.5) +
  facet_grid(~method,scales="free")+ see::scale_color_flat()+theme_light()    + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position="bottom")    
```                   
