---
title: "Systematic review and meta-analysis of Real-World Evidence"
authors:   
  - name: Dimitris Mavridis
    affiliations:
      - ref: ioannina
  - name: Thomas Debray
    orcid: 0000-0002-1790-2719
    affiliations:
      - ref: smartdas
affiliations:
  - id: smartdas
    name: Smart Data Analysis and Statistics B.V.
    city: Utrecht
  - id: ioannina
    name: University of Ioannina
    city: Ioannina
format:
  html:
    toc: true
    number-sections: true
execute:
  cache: true
bibliography: 'https://api.citedrive.com/bib/0d25b38b-db8f-43c4-b934-f4e2f3bd655a/references.bib?x=eyJpZCI6ICIwZDI1YjM4Yi1kYjhmLTQzYzQtYjkzNC1mNGUyZjNiZDY1NWEiLCAidXNlciI6ICIyNTA2IiwgInNpZ25hdHVyZSI6ICI0MGFkYjZhMzYyYWE5Y2U0MjQ2NWE2ZTQzNjlhMWY3NTk5MzhhNzUxZDNjYWIxNDlmYjM4NDgwOTYzMzY5YzFlIn0=/bibliography.bib'
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(kableExtra)
```

## Introduction

We first load the required packages

```{r}
#| warning: false
library(dplyr)
library(netmeta)
```

```{r}
#| echo: false
#| message: false
#| warning: false
library(dmetar)
```

## Pairwise meta-analysis of clinical trials

### Toculizumab for coronavirus disease 2019
In this example, we consider the results from a systematic literature review of clinical trials investigating any pharmacological in hosptialized patients with coronavirus disease 2019 [@selvarajan_efficacy_2022]. A total of 23 randomized controlled trials were included and studied seven different interventions: dexamethasone, remdesivir, tocilizumab, hydroxychloroquine, combination of lopinavir/ritonavir, favipiravir and interferon-β. We here focus on the synthesis of 7 trials that comparted toculizumab (`TOCI`) to standard care (`STD`)
and collected mortality data. 

```{r}
#| echo: false
#| message: false
#| warning: false
studlab <- c(rep("Tomazini et al",2), rep("Horby et al",2), 
             rep("Biegel et al",2), rep("Wang et al",2),
             rep("Spinner et al",2),rep("Solidarity trial - Remdesivir",2),
             rep("Hermine et al",2),rep("Salvarini et al",2),
             rep("Stone et al",2), rep("Rosas et al",2), rep("Salama et al",2),
             rep("Veiga et al",2), rep("Ulrich et al",2),
             rep("Lyngbakken et al",2), rep("Solidarity trial - HCQ",2), 
             rep("Abd-Elsalam et al",2),rep("Cavalcanti et al",2),
             rep("Recovery group",2), rep("Solidarity trial-LR",2), 
             rep("Cao et al",2), rep("Davoudi-Monfared",2), 
             rep("Solidarity trial - Interferon",2),rep("Rohmani",2))

treat <- c("DEXA", "STD", "DEXA", "STD", "REM", "STD", "REM", "STD", "REM", 
           "STD", "REM", "STD", "TOCI", "STD", "TOCI", "STD", "TOCI", "STD",
           "TOCI", "STD", "TOCI", "STD", "TOCI", "STD", "HCQ", "STD", "HCQ", 
           "STD", "HCQ", "STD", "HCQ", "STD", "HCQ", "STD", "LOP+RIT",  
           "STD", "LOP+RIT", "STD", "LOP+RIT", "STD", "INT-B", "STD", "INT-B",
           "STD", "INT-B", "STD")    

event <- c(85,91,482,1110,59,77,22,10,5,4,301,303,7,8,2,1,9,3,58,28,26,11,14,6,13,6,1,1,104,84,6,5,1,1,374,767,148,146,19,25,8,15,243,216,2,6)
n <- c(151,148,2104,4321,541,521,158,78,396,200,2743,2708,63,67,60,66,161,82,294,144,249,128,65,64,67,61,27,26,947,906,97,97,221,227,1616,3424,1399,1372,99,100,42,39,2050,2050,33,33)

covid <- pairwise(treat = treat, event = event, n = n, studlab = studlab, sm = "OR")

tocilizumab <- covid %>% filter(treat1 == "TOCI")
remdesvir <- covid %>% filter(treat1 == "REM")
```


```{r}
#| echo: false
#| message: false
#| warning: false


tocilizumab %>% select(studlab, treat1, treat2, event1, n1, event2, n2) %>% 
  kable(
      align = "lrrrrrr",
      booktabs = TRUE,
      longtable = TRUE,
      linesep = ""
    ) %>%
    kableExtra::kable_styling(
      latex_options = c("repeat_header"),
    )
```

We now conduct a pairwise meta-analysis to assess the pooled effect of tocilizumab versus standard care. For each study, the log odds ratio and corresponding standard error is derived after which the corresponding estimates are pooled using the  Mantel-Haenszel method.

```{r}
#| fig-width: 6
#| fig-height: 5
results.TOCI <- metabin(event1,n1,event2,n2,studlab,data=tocilizumab,
                        sm="OR",main="tocilizumab vs standard care", 
                        prediction=TRUE)
forest(results.TOCI, leftcols = "studlab", rightcols = "effect.ci")
```

Altough a random effects meta-analysis was conducted, no heterogeneity was found ($\tau$=`r results.TOCI$tau`, with a 95\% confidence interval ranging from `r  results.TOCI$lower.tau` to `r round(results.TOCI$upper.tau,2)`).


### Remdesivir for coronavirus disease 2019
In aforementioned example, a total of 4 trials compared remdesivir to standard care:

```{r}
#| echo: false
#| fig-width: 6
results.REM <- metabin(event1,n1,event2,n2,studlab,data=remdesvir,
                        sm="OR",main="remdesvir vs standard care", 
                        prediction=TRUE)
forest(results.REM, leftcols = "studlab", rightcols = "effect.ci")
```


## Network meta-analysis of clinical trials

### Interventions for coronavirus disease 2019
We now consider a larger network of evidence presented by Sandhiya et al [@selvarajan_efficacy_2022]. The following interventions were investigated:

* Dexamethasone (`DEXA`)
* Favipravir (``)
* Hydroxychloroquine (`HCQ`)
* Interferon β (`INT-B`)
* Lopinavir + ritonavir (`LOP+RIT`)
* Remdesivir 200/100 mg for 5 or 10 days (`REM`)
* Tocilizumab (`TOCI`)

The control treatment is standard care (`STD`).

```{r}
#| echo: false
#| message: false
#| warning: false

covid %>% select(studlab, treat1, treat2, event1, n1, event2, n2) %>% 
  kable(
      align = "lrrrrrr",
      booktabs = TRUE,
      longtable = TRUE,
      linesep = ""
    ) %>%
    kableExtra::kable_styling(
      latex_options = c("repeat_header"),
    )
```

The corresponding network is displayed below:

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-cap: Evidence network of the 23 coronavirus-19 trials

NMA.covid <- netmeta(TE = TE, seTE = seTE, treat1 = treat1, treat2 = treat2,
                     studlab = studlab, data = covid, sm = "OR", ref = "STD",
                     comb.random = TRUE)

netgraph(NMA.covid)
```

Below, we conduct a random effects network meta-analysis where we consider 
standard care (`treat = "STD"`) as the control treatment.

```{r}
NMA.covid <- netmeta(TE = TE, seTE = seTE, treat1 = treat1, treat2 = treat2,
                     studlab = studlab, data = covid, sm = "OR", ref = "STD",
                     comb.random = TRUE)
forest(NMA.covid, leftcols = "studlab", rightcols = "effect.ci")
```

The figure below shows the percentage of direct and indirect evidence used for each estimated comparison. 

```{r}
#| echo: false
#| fig-width: 10
d.evidence <- direct.evidence.plot(NMA.covid)
plot(d.evidence)
```

A more detailed tutorial on netmeta is available from the book [Doing Meta-Analysis with R: A Hands-On Guide](https://bookdown.org/MathiasHarrer/Doing_Meta_Analysis_in_R/).


### Pharmacologic treatments for chronic obstructive pulmonary disease 
In this example, we consider the resuls from a systematic review of randomized controlled trials on pharmacologic treatments for chronic obstructive pulmonary disease [@Baker_2009]. The primary outcome, occurrence of one or more episodes of COPD exacerbation, is binary (yes / no). For this outcome, five drug treatments (fluticasone, budesonide, salmeterol, formoterol, tiotropium) and two combinations (fluticasone + salmeterol, budesonide + formoterol) were compared to placebo. The authors considered the two combinations as separate treatments instead of evaluating the individual components.

```{r}
data(Baker2009)
```

```{r}
#| echo: false
kable(head(Baker2009))
```

```{r}
Baker <- pairwise(treat = treatment,
                  event = exac,
                  n = total,
                  studlab = id,
                  sm = "OR",
                  data = Baker2009)

NMA.COPD <- netmeta(TE = TE, seTE = seTE, treat1 = treat1, treat2 = treat2,
                    studlab = studlab, data = Baker, sm="OR", ref = "Placebo",
                    comb.random = TRUE)

netgraph(NMA.COPD)

```


### Advanced Therapies for Ulcerative Colitis
In this example, we consider a systematic literature review of Phase 3 
randomized controlled trials investigating the following advanced therapies: 
infliximab, adalimumab, vedolizumab, golimumab, tofacitinib, ustekinumab, 
filgotinib, ozanimod, and upadacitinib [@panaccione_efficacy_2023]. This review 
included 48 RCTs, from which 23 were found eligible for inclusion in a 
network meta-analysis. The included RCT populations were largely comparable in their baseline
characteristics, though some heterogeneity was noted in weight, 
disease duration, extent of disease, and concomitant medications. 
A risk of bias assessment showed a low risk of bias for all included RCTs, 
which were all industry sponsored.

We here focus on the synthesis of 18 trials that contributed efficacy data for 
induction in bio-naive populations. The following FDA- and/or EMA-approved 
biologic or SMD doses were investigated: 

* Adalimumab subcutaneous 160 mg at week 0, 80 mg at week 2, and 40 mg at week 4 (`ADA160/80`)
* Infliximab intravenous 5 mg/kg (`INF5`) at weeks 0, 2, and 6 then every 8 weeks
* Infliximab intravenous 10 mg/kg (`INF10`) at weeks 0, 2, and 6 then every 8 weeks
* Filgotinib oral 100 mg once daily (`FIL100`)
* Filgotinib oral 200 mg once daily (`FIL200`)
* Golimumab subcutaneous 200 mg at week 0 and 100 mg at week 2 (`GOL200/100`)
* Ozanimod oral 0.23 mg once daily for 4 days, 0.46 mg once daily for 3 days, then 0.92 mg once daily (`OZA0.92`)
* Tofacitinib oral 10 mg twice daily for 8 weeks (`TOF10`)
* Upadacitinib oral 45 mg once daily for 8 weeks (`UPA45`)
* Ustekinumab intravenous 6 mg/kg at week 0 (`UST6`)
* Vedolizumab intravenous 300 mg at weeks 0, 2, and 6 (`VED300`)

The reference treatment is placebo (`PBO`).


```{r}
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: Efficacy outcomes (i.e., clinical remission) data of induction bio-naïve populations
# Efficacy outcomes data of induction bio-naïve populations
studlab <- c("ACT-1", "ACT-1", "ACT-1", "ACT-2", "ACT-2", "ACT-2", "GEMINI 1",
             "GEMINI 1", "Japic CTI-060298", "Japic CTI-060298", "Jiang 2015", 
             "Jiang 2015", "M10-447", "M10-447", "NCT01551290", "NCT01551290",
             "NCT02039505", "NCT02039505", "OCTAVE 1", "OCTAVE 1", "OCTAVE 2",
             "OCTAVE 2", "PURSUIT-SC", "PURSUIT-SC", "SELECTION", "SELECTION",
             "SELECTION", "TRUE NORTH", "TRUE NORTH", "U-ACCOMPLISH", 
             "U-ACCOMPLISH", "U-ACHIEVE Study 2", "U-ACHIEVE Study 2", 
             "ULTRA-1", "ULTRA-1", "ULTRA-2", "ULTRA-2", "UNIFI", "UNIFI")

treat <- c("INF10", "INF5", "PBO", "INF10", "INF5", "PBO", "VED300", "PBO", 
           "INF5", "PBO", "INF5", "PBO", "ADA160/80", "PBO", "INF5", "PBO", 
           "VED300", "PBO", "TOF10", "PBO", "TOF10", "PBO", "GOL200/100", 
           "PBO", "FIL100", "FIL200", "PBO", "OZA0.92", "PBO", "UPA45", "PBO",
           "UPA45", "PBO", "ADA160/80", "PBO", "ADA160/80", "PBO", "UST6",
           "PBO")


event <- c(39, 47, 18, 33, 41, 7, 30, 5, 21, 11, 22, 9, 9, 11, 11, 5, 22,
           6, 56, 9, 43, 4, 45, 16, 47, 60, 17, 66, 10, 54, 3, 41, 4, 24,
           12, 32, 16, 27, 15)

n <- c(122, 121, 121, 120, 121, 123, 130, 76, 104, 104, 41, 41, 90, 96, 50,
       49, 79, 41, 222, 57, 195, 47, 253, 251, 277, 245, 137, 299, 151, 166,
       81, 145, 72, 130, 130, 150, 145, 147, 151)

UlcerativeColitis <- pairwise(treat=treat,event=event,n=n,studlab=studlab, sm="OR")

UlcerativeColitis %>% select(studlab, treat1, treat2, event1, n1, event2, n2) %>% 
  kable(
      align = "lrrrrrr",
      booktabs = TRUE,
      longtable = TRUE,
      linesep = ""
    ) %>%
    kableExtra::kable_styling(
      latex_options = c("repeat_header"),
    )
```

The corresponding network is displayed below:

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-cap: Evidence network of 18 trials that contributed efficacy data for induction in bio-naive populations

NMA.uc <- netmeta(TE = TE, seTE = seTE, treat1 = treat1, treat2 = treat2,
                  studlab = studlab, data = UlcerativeColitis, sm = "OR", 
                  ref = "PBO", comb.random = TRUE)

netgraph(NMA.uc)
```

Below, we conduct a random effects network meta-analysis of the reported study 
effects (expressed as odds ratio) and consider placebo (`treat = "PBO"`) as 
the control treatment.

```{r}
#| eval: false
NMA.uc <- netmeta(TE = TE, seTE = seTE, treat1 = treat1, treat2 = treat2,
                  studlab = studlab, data = UlcerativeColitis, sm = "OR", 
                  ref = "PBO", comb.random = TRUE)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 6
forest(NMA.uc, leftcols = "studlab", rightcols = "effect.ci")
```

All treatments except `FIL100` and `UST6` are significantly more efficacious than `PBO` at inducing clinical remission. We can now estimate the probabilities of each treatment being at each possible rank and the SUCRAs (Surface Under the Cumulative RAnking curve):

```{r}
sucra.uc <- rankogram(NMA.uc, nsim = 100, random = TRUE, common = FALSE, 
                      small.values = "undesirable")

# Exctract the SUCRA values
sucra.uc$ranking.random
```

These results indicate that `r round( sucra.uc$ranking.random["UPA45"]*100,1)`% of the evaluated treatments are worse than `UPA45`.





## Version info {.unnumbered}
This chapter was developed using the following version of R and its packages:

```{r}
#| echo: false
#| message: false
#| warning: false
sessionInfo()
```


